---
title: "THE SWEDISH PHARMACY RESIDENT RESEARCH PROJECT"
author: 
  - Jack B. Huber, Ph.D.
  - Senior Clinical Data Analyst
date: "`r format(Sys.Date(), '%B %Y')`"
documentclass: report
bibliography: [bibfile.bib,packages.bib]
link-citations: yes
output:
  html_document:
    theme: cerulean
    # toc: yes
    # number_sections: yes
    # toc_float:
    #   collapsed: yes
    #   smooth_scroll: yes
    citation_package: natbib
    fig_caption: yes
    # includes:
      # in_header: logo.html
---

```{r eval=FALSE, echo=FALSE}
  prettydoc::html_pretty:
    theme: hpstr
  pdf_document:
    includes:
      in_header: preamble.tex
    latex_engine: xelatex
    citation_package: natbib
    keep_tex: yes
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::write_bib(c(.packages(),"knitr","pwr","rmarkdown","VIM"),"packages.bib")
```

---

# Purpose of this Site

The purpose of this site^[This site was built in `R` [@R-base] with `R Markdown` [@R-rmarkdown].] is to offer targeted advice and resources to support Swedish Pharmacy residents in their journey from project proposal to a completed research project.

The menu above illustrates the organization of the resident research project and of this site. Like most research projects, the resident project goes through a sequence of phases or stages beginning with the planning and design of the project; evolving through middle stages of collecting, organizing, managing, and analyzing the data; and culminating in reporting the data in tables and graphs for manuscripts and visual presentation.

Each page attempts to identify and deal with important issues that come up. This includes explanation, advice, and links to resources where available and appropriate.

These pages are works in progress. The intent is for them to evolve in adaptation to the needs of this year's residents. I will continue to edit them as I see fit; and if something is missing, please don't hesitate to reach out to me: [jack.huber@swedish.org](mailto:jack.huber@swedish.org).^[A little bit [about me](AboutMe.html)[.](Projects.html)]

---
`r if (knitr::is_html_output()) '
# References {-}
'`
<div id="refs"></div>